---
title: "Simulation Study on Cluster Randomized Control Design"
author: "Yunan Chen"
date: "2024-12-02"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lmerTest)
library(lme4)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(gt)
library(RColorBrewer)
```

# Abstract

Cluster randomized trials are widely utilized in public health and clinical research to evaluate treatment effects, requiring careful design to balance the number of clusters, measurements per cluster, and associated costs under budget constraints. This study employs simulation studies guided by the ADEMP framework to optimize resource allocation strategies for minimizing the variance of treatment effect estimates ($\hat{\beta}$). Using Normal and Poisson data-generating mechanisms, we examine the impacts of cost ratios, variance components of the distributions, and treatment effect sizes on the precision and power of $\hat{\beta}$. Results demonstrate that increasing the number of clusters reduces variance but with diminishing returns, emphasizing the need for balance in resource allocation. Poisson outcomes exhibit greater sensitivity to between-cluster variability compared to Normal outcomes, reflecting differences in the hierarchical data structures and their influence on intra-class correlation (ICC). Higher ICC values reduce the effective sample size by increasing within-cluster correlation, inflating $\hat{\beta}$ variance. Conversely, lower ICC values improve the precision of $\hat{\beta}$ by enhancing the contribution of within-cluster data. These findings underscore the importance of tailoring resource allocation strategies to specific data distributions and ICC properties to design cost-efficient and statistically robust cluster randomized trials.

# Introducation

Cluster randomized trials are widely used in public health and clinical
research, where clusters of units are randomized to either control or
treatment groups to estimate treatment effects. For example, in a study
evaluating the effectiveness of a new hand hygiene protocol, hospitals
were randomized to either implement the protocol or continue standard
practices. Within each hospital, individual patient outcomes, such as
rates of hospital-acquired infections, were measured to assess the
impact of the intervention (Clancy et al., 2020). In another study,
individuals were treated as clusters, and repeated measurements, such as
daily blood glucose levels, were collected to assess the effect of a
dietary intervention on long-term glycemic control (Karimian et al.,
2023).

In our study, we consider clusters as individuals and the number of
measurements per cluster as repeated measurements of these individuals. A critical design challenge for cluster randomized trials
lies in allocating a fixed budget to maximize the precision of treatment
effect estimates. This challenge is further complicated by uncertainties
in intracluster correlation coefficients, which measure within-cluster
similarity. While independent measurements are generally preferred for
statistical power, budget constraints often necessitate strategic
trade-offs between the number of clusters and the number of measurements
per cluster. In some cases, obtaining additional measurements within the
same cluster or subject can be significantly less expensive than
recruiting new clusters or subjects. For example, with a fixed budget,
sampling a smaller number of subjects with repeated measurements may
yield comparable or even improved precision for estimating treatment
effects by reducing residual variance.

This project, in collaboration with Dr. Zhijin Wu from the Biostatistics
Department, aims to develop a simulation study to identify optimal
cluster randomized trial designs under budgetary constraints. Using
simulation studies guided by the aim, data structure, estimand, methods,
and performance (ADEMP) framework, the study systematically explores
trade-offs between cluster-level and within-cluster measurements under
varying cost structures and statistical assumptions. By evaluating
resource allocation strategies, the project provides practical
recommendations for designing cost-effective and statistically robust
cluster randomized trials that maximize the precision of treatment
effect estimates. Additionally, extending the simulations to outcomes
modeled with a Poisson distribution will offer insights into how
hierarchical modeling assumptions impact design recommendations,
contributing to advancing the methodology for resource-efficient
experimental designs in cluster randomized trials.

# Method

To achieve the aims of the project, we are employing the ADEMP
framework, a structured approach for designing and reporting simulation
studies, ensuring clarity, reproducibility, and thorough evaluation of
statistical methods. The framework includes five key elements: Aims,
which specify the study's objectives, such as evaluating bias,
precision, or power; Data-generating mechanisms, which define how data
are simulated, including underlying models and parameter values;
Estimands, identifying the quantities or parameters of interest being
assessed; Methods, describing the statistical techniques or models
applied to estimate these quantities; and Performance measures, which
evaluate the methods' effectiveness using metrics like bias, mean
squared error, variance or power. This systematic approach guides the
design and analysis of simulation studies to achieve robust and
meaningful insights. By adopting this structured approach, the framework
ensures methodological rigor, enhances reproducibility, and provides a
comprehensive evaluation of statistical methods within the context of
simulation-based research.

## Simulation framework

**Aims:**

To evaluate the impacts (1) of the optimal allocation of a fixed budget across the number of clusters ($G$) and the number of measurements per cluster ($R$) in a cluster randomized trial, focusing on minimizing the variance of the estimated treatment effect ($\beta$) while balancing trade-offs between within-cluster and between-cluster variability; (2) of underlying data-generating parameters, such as $\beta$ and $\sigma^2$ for Normal outcomes and $\beta$ and $\gamma^2$ for Poisson outcomes, as well as relative cost ratios ($c_1/c_2$), on the precision and power of the treatment effect estimate; and (3) of differences in performance under Normal and Poisson data-generating mechanisms, investigating how distributional assumptions influence the results and optimal design strategies.


**Data-Generating Mechanisms:** 

- Normal Outcomes: For measurement $j = 1, \cdots, R$ within cluster $i = 1, \cdots, G$, generate the data under the hierarchical structure:
\begin{align*}
        \mu_{i0} &= \alpha + \beta X_i \quad \text{where $X_i = 0$ for control and $X_i = 1$ for treatment}, \\
        \mu_i \mid \epsilon_i &= \mu_{i0} + \epsilon_i, \quad \epsilon_i \sim N(0, \gamma^2), \\
        Y_{ij} \mid \mu_i &= \mu_i + e_{ij}, \quad e_{ij} \sim N(0, \sigma^2).
    \end{align*} 
    
-   Poisson Outcomes: For measurement $j = 1, \cdots, R$ within cluster $i = 1, \cdots, G$, let:
    \begin{align*}
          \log(\mu_i) &\sim N(\alpha + \beta X_i, \gamma^2), \quad \text{where $X_i = 0$ for control and $X_i = 1$ for treatment}, \\
          Y_{ij} \mid \mu_i &\sim \text{Poisson}(\mu_i)
      \end{align*}

**Estimand:** 

The primary estimand is the treatment effect, $\beta$,
which quantifies the difference in outcomes between the treatment
($X_i = 1$) and control ($X_i = 0$) groups. For Normal outcomes, $\beta$
represents the average difference in means. For Poisson outcomes,
$\beta$ corresponds to the log-risk ratio of the outcome rates.

**Methods:** 

For each simulation, the generated data set ($X$, $Y$) under normal or Poisson data-generating mechanisms is analyzed using a GLM model, and the estimate $\beta$ is then extracted from the model result.

**Performance Measures:** 

We will assess the performance of the study design using two key metrics: variance and power. Variance measures the variability of $\hat{\beta}$ across simulations, providing insights into how design parameters, such as the number of clusters and measurements per cluster, impact the precision of the treatment effect estimate. Power quantifies the ability to detect significant treatment effects by calculating the proportion of simulations where $\hat{\beta}$ is significantly different from zero. This evaluation focuses on the influence of cost ratios and variability parameters on the statistical power of the study design.

## Simulation process

The simulation process involved two major steps. First, we fixed the
key parameters of the distribution:
$\alpha=2, \beta=1.5, \sigma^2=1, \gamma^2=1$ These values were chosen
because they represent moderate, plausible magnitudes for intercepts,
effect sizes, and variances in typical clustered designs, allowing us to
explore the study design under realistic and interpretable baseline
conditions. With these baseline values and a fixed budget of 2000, we
systematically varied the relative cost $c_1/c_2$ across 2, 5,
10, and 20, and the number of clusters over a sequence from 10 to 50 in
increments of 5. To compute the cost per measurement $c_2$, we used
$c_2 = \frac{c_1}{\text{relative cost}}$. The number of measurements per
cluster was then calculated as
$\text{number of measurements per Cluster} = \frac{\text{Budget}-n_{clusters}\cdot c_1}{n_{clusters}\cdot c_2}$.
For example, under a fixed budget of 2000, if the relative cost
c_1/c_2 is 2 and there are 20 clusters, the number of
measurements per cluster would be 98. However, if the relative cost
increases to 10 while keeping the number of clusters at 20, the number
of measurements per cluster decreases to 90. This demonstrates that
higher relative cluster costs or an increase in the number of clusters
reduces the number of measurements per cluster, as the budget must be
distributed accordingly. The goal of this step was to identify the
optimal combination of the number of clusters and the number of
measurements per cluster that minimizes the variance of the estimated
treatment effect $\beta$.

After determining the optimal combination of the number of clusters and
the number of measurements per cluster, we fixed these values and
conducted additional simulations to assess how varying the distribution
parameters influences the optimal study design. Specifically, we varied
the residual variance $\sigma^2$ over a sequence from 0.01 to 10 with 10
equally spaced values, and the treatment effect $\beta$ across 0.05,
0.5, and 1.5. Changes in $\sigma^2$ directly affect the intra-class
correlation (ICC), which quantifies the proportion of the total variance
attributable to clustering effects, given by
$\text{ICC} = \frac{\gamma^2}{\gamma^2+\sigma^2}$. Under a fixed
number of clusters and measurements per cluster, increasing $\sigma^2$
reduces the ICC, indicating weaker clustering effects. This, in turn,
increases the within-cluster variation relative to the total variation,
which can result in greater variability of the $\beta$ estimate. As
$\sigma^2$ increases, the residual noise dominates, making it harder to
detect the treatment effect precisely and potentially widening
confidence intervals for $\beta$. Conversely, lower values of $\sigma^2$
strengthen clustering effects, reducing the variability of $\beta$ and
improving the precision of the estimate. The parameter $\alpha$,
representing the intercept, was not varied because it does not influence
the variance or clustering structure of the data. Instead, $\alpha$
shifts the overall distribution of outcomes without altering the
relationships between the variables or the precision of the estimates,
and therefore has no direct impact on the performance metrics of the
study design. This step allowed us to evaluate not only the robustness
of the optimal design but also how the degree of residual variance
impacts the precision of $\beta$ under different data generation
conditions. By combining these two steps, we were able to
comprehensively evaluate and refine the study design to achieve a
balance between cost efficiency and statistical precision.

# Results

## Optimal Q and R

```{r warning=FALSE, message=FALSE}
# Simulate data and performance matrix under Normal distribution
sim_normal <- function(n_clusters, B, c1, c1_c2_ratio, alpha, beta, gamma2, sigma2, n_sim, alpha_level = 0.05) {
  set.seed(2550)
  # Calculate c2 from the ratio
  c2 <- c1 / c1_c2_ratio
  
  # Calculate the number of measurements per cluster
  n_obs_per_cluster <- floor((B - n_clusters * c1) / (n_clusters * c2)) + 1
  
  # Assign clusters to treatment (X = 1) or control (X = 0)
  cluster_treatment <- rep(c(0, 1), n_clusters / 2) 
  
  # Initialize data frame for metrics
  all_metrics <- data.frame(
    beta_est = numeric(n_sim),
    beta_bias = numeric(n_sim),
    power = numeric(n_sim),
    coverage = numeric(n_sim)
  )
  
  for (sim in 1:n_sim) {
    # Generate random cluster-level effects
    cluster_effects <- rnorm(n_clusters, mean = 0, sd = sqrt(gamma2))
    # Generate cluster-level means
    cluster_means <- alpha + beta * cluster_treatment + cluster_effects
    
    # Simulate data
    cluster_data <- data.frame(
      cluster_id = rep(1:n_clusters, each = n_obs_per_cluster)[1:(n_clusters * n_obs_per_cluster)],
      X = rep(cluster_treatment, each = n_obs_per_cluster)[1:(n_clusters * n_obs_per_cluster)],
      Y = rnorm(
        n_clusters * n_obs_per_cluster,
        mean = cluster_means[rep(1:n_clusters, each = n_obs_per_cluster)],
        sd = sqrt(sigma2)
      )[1:(n_clusters * n_obs_per_cluster)]
    )
    
    # Fit a linear mixed-effects model
    model <- lmerTest::lmer(Y ~ X + (1 | cluster_id), data = cluster_data)
    
    # Extract measurements
    beta_est <- fixef(model)["X"]
    ci <- confint(model, parm = "X", method = "Wald")
    ci_coverage <- (beta >= ci[1] & beta <= ci[2])
    p_value <- summary(model)$coefficients["X", "Pr(>|t|)"]
    
    # Store simulation results
    all_metrics[sim, ] <- c(
      beta_est = beta_est,
      beta_bias = beta_est - beta,
      power = ifelse(p_value < alpha_level, 1, 0),
      coverage = ci_coverage
    )
  }
  
  # Compute performance metrics
  beta_est_var <- var(all_metrics$beta_est)
  min_beta_est <- min(all_metrics$beta_est)
  max_beta_est <- max(all_metrics$beta_est)
  avg_metrics <- colMeans(all_metrics)
  
  # Return results
  results <- data.frame(
    n_clusters = n_clusters,
    n_obs_per_cluster = n_obs_per_cluster,
    true_beta = beta,
    beta_est_mean = avg_metrics["beta_est"],
    min_beta_est = min_beta_est,
    max_beta_est = max_beta_est,
    beta_bias_mean = avg_metrics["beta_bias"],
    beta_est_var = beta_est_var,
    power = avg_metrics["power"],
    ci_coverage = avg_metrics["coverage"]
  )
  
  return(list(
    metrics = results,
    simulated_data = cluster_data
  ))
}

```

```{r warning=FALSE, message=FALSE}
sim_poisson <- function(n_clusters, B, c1, c1_c2_ratio, alpha, beta, gamma2, n_sim, alpha_level = 0.05) {
  set.seed(2550)
  # calculate c2 from the ratio
  c2 <- c1 / c1_c2_ratio
  
  # calculate the number of measurements per cluster
  n_obs_per_cluster <- floor((B - n_clusters * c1) / (n_clusters * c2)) + 1

  
  # assign clusters to treatment (X = 1) or control (X = 0)
  cluster_treatment <- rep(c(0,1), n_clusters/2) 
  
  # initialize data frame
  all_metrics <- data.frame(
    beta_est = numeric(n_sim),
    beta_bias = numeric(n_sim),
    power = numeric(n_sim),
    coverage = numeric(n_sim)
  )
  
  for (sim in 1:n_sim) {
    
    # generate random cluster-level effects (log scale)
    cluster_effects <- rnorm(n_clusters, mean = alpha, sd = sqrt(gamma2))
    # generate cluster-level means (log scale)
    log_mu <- alpha + beta * cluster_treatment + cluster_effects
    mu <- exp(log_mu)  
    
  cluster_data <- data.frame(
  cluster_id = rep(1:n_clusters, each = n_obs_per_cluster)[1:(n_clusters * n_obs_per_cluster)],
  X = rep(cluster_treatment, each = n_obs_per_cluster)[1:(n_clusters * n_obs_per_cluster)],
  Y = rpois(n_clusters * n_obs_per_cluster, lambda = mu[rep(1:n_clusters, each = n_obs_per_cluster)])[1:(n_clusters * n_obs_per_cluster)])
    
    # fit a Poisson GLM with random intercept for clusters
    model <- lme4::glmer(Y ~ X + (1 | cluster_id), data = cluster_data, family = poisson())
    
    # extract measurements
    beta_est <- fixef(model)["X"]
    ci <- confint(model, parm = "X", method = "Wald")
    ci_coverage <- (beta >= ci[1] & beta <= ci[2]) 
    p_value <- summary(model)$coefficients["X", "Pr(>|z|)"]
    
    # store simulation results
    all_metrics[sim, ] <- c(
      beta_est = beta_est,
      beta_bias = beta_est - beta,
      power = ifelse(p_value < alpha_level, 1, 0),
      coverage = ci_coverage
    )
  }
  
  # Compute performance 
  beta_est_var <- var(all_metrics$beta_est)
  min_beta_est <- min(all_metrics$beta_est)
  max_beta_est <- max(all_metrics$beta_est)
  avg_metrics <- colMeans(all_metrics)
  
  # Return results
  results <- data.frame(
    n_clusters = n_clusters,
    n_obs_per_cluster = n_obs_per_cluster,
    true_beta = beta,
    beta_est_mean = avg_metrics["beta_est"],
    min_beta_est = min_beta_est,
    max_beta_est = max_beta_est,
    beta_bias_mean = avg_metrics["beta_bias"],
    beta_est_var = beta_est_var,
    power = avg_metrics["power"],
    ci_coverage = avg_metrics["coverage"]
  )
  
  return(list(
    metrics = results,
    simulated_data = cluster_data
  ))
}
```

```{r warning=FALSE, message=FALSE}
# Function to find optimal # of clusters and # of obs/cluster under Normal
sim_normal_opt <- function(n_clusters_seq, c1_c2_ratios, B, c1, alpha, beta, sigma2, gamma2, n_sim, alpha_level = 0.05) {
  
  # Initialize storage for all results
  all_results <- list()
  
  # Loop over parameters (# of clusters and cost ratio)
  for (n_clusters in n_clusters_seq) {
    for (c1_c2_ratio in c1_c2_ratios) {
      # Run simulation for current parameter set
      result <- sim_normal(
        n_clusters = n_clusters, 
        B = B, 
        c1 = c1, 
        c1_c2_ratio = c1_c2_ratio, 
        alpha = alpha, 
        beta = beta, 
        gamma2 = gamma2, 
        sigma2 = sigma2, 
        n_sim = n_sim, 
        alpha_level = alpha_level
      )
      
      # Store the results
      all_results <- append(all_results, list(
        result$metrics %>% mutate(c1_c2_ratio = c1_c2_ratio)
      ))
    }
  }
  
  # Combine all results into a single data frame
  combined_results <- bind_rows(all_results)
  return(combined_results)
}

# Set Parameters
n_clusters_seq <- seq(10, 50, 5)
c1_c2_ratios <- c(2, 5, 10, 20)

B <- 2000
c1 <- 20
alpha <- 2
beta <- 1.5
sigma2 <- 1
gamma2 <- 1
n_sim <- 100

# Run simulations over the grid
# res_normal_opt <- sim_normal_opt(
#   n_clusters_seq = n_clusters_seq,
#   c1_c2_ratios = c1_c2_ratios,
#   B = B,
#   c1 = c1,
#   alpha = alpha,
#   beta = beta,
#   sigma2 = sigma2,
#   gamma2 = gamma2,
#   n_sim = n_sim
# )
# # Store as .csv file
# write.csv(res_normal_opt, "res_normal_opt.csv")
res_normal_opt <- read.csv("res_normal_opt.csv")
```

```{r warning=FALSE, message=FALSE}
# Function to find optimal # of clusters and # of obs/cluster under Poisson
sim_poisson_opt <- function(n_clusters_seq, c1_c2_ratios, B, c1, alpha, beta, gamma2, n_sim, alpha_level) {
  all_results <- list()
  
  # Loop over the parameter grid
  for (c1_c2_ratio in c1_c2_ratios) {
    for (n_clusters in n_clusters_seq) {
      # Run the simulation for each configuration
      result <- sim_poisson(
        n_clusters = n_clusters,
        B = B,
        c1 = c1,
        c1_c2_ratio = c1_c2_ratio,
        alpha = alpha,
        beta = beta,
        gamma2 = gamma2,
        n_sim = n_sim,
        alpha_level = alpha_level
      )
      
      # Store the results
      all_results <- append(all_results, list(
        result$metrics %>% mutate(c1_c2_ratio = c1_c2_ratio)
      ))
    }
  }
  
  # Combine all results into a single data frame
  combined_results <- bind_rows(all_results)
  return(combined_results)
}

# Set Parameters
n_clusters_seq <- seq(10, 50, 5)  
c1_c2_ratios <- c(2, 5, 10, 20) 

# Run the evaluation
# res_poisson_opt <- sim_poisson_opt(
#   n_clusters_seq = n_clusters_seq,
#   c1_c2_ratios = c1_c2_ratios,
#   B = 2000,
#   c1 = 20,
#   alpha = 2,
#   beta = 1.5,
#   gamma2 = 1,
#   n_sim = 100,
#   alpha_level = 0.05
# )
# # Store as .csv file
# write.csv(res_poisson_opt, "res_poisson_opt.csv")
res_poisson_opt <- read.csv("res_poisson_opt.csv")
```


`Table 1` summarizes the variance of the treatment effect estimate ($\hat{\beta}$) and the corresponding number of measurements per cluster ($R$) across various combinations of the number of clusters ($G$) and relative cost ratios ($c_1/c_2$) for both Normal and Poisson data-generating mechanisms. Each row represents a specific number of clusters, while the columns provide the corresponding $R$ and the variance of $\hat{\beta}$ for cost ratios $c_1/c_2 = 2, 5, 10, 20$, under a fixed budget of 2000. The highlighted cells identify the optimal combinations of $G$ and $R$ that achieve the lowest variance of $\hat{\beta}$, reflecting the most efficient allocation of resources within the given constraints.


```{r warning=FALSE, message=FALSE, out.width="80%"}
# Results Table (Normal)
r2_df <- res_normal_opt %>% 
  filter(c1_c2_ratio == 2) %>% 
  mutate(beta_est_var = round(beta_est_var, 2)) %>%
  select(`G` = n_clusters, `# of Obs / Cluster (2)` = n_obs_per_cluster, `Var(beta_est) (2)` = beta_est_var)

r5_df <- res_normal_opt %>% 
  filter(c1_c2_ratio == 5) %>% 
  mutate(beta_est_var = round(beta_est_var, 2)) %>%
  select(`# of Obs / Cluster (5)` = n_obs_per_cluster, `Var(beta_est) (5)` = beta_est_var)

r10_df <- res_normal_opt %>% 
  filter(c1_c2_ratio == 10) %>% 
  mutate(beta_est_var = round(beta_est_var, 2)) %>%
  select(`# of Obs / Cluster (10)` = n_obs_per_cluster, `Var(beta_est) (10)` = beta_est_var)

r20_df <- res_normal_opt %>% 
  filter(c1_c2_ratio == 20) %>% 
  mutate(beta_est_var = round(beta_est_var, 2)) %>%
  select(`# of Obs / Cluster (20)` = n_obs_per_cluster, `Var(beta_est) (20)` = beta_est_var)

combined_df <- bind_cols(r2_df, r5_df, r10_df, r20_df)

# Results Table (Poisson)
r2_df_poisson <- res_poisson_opt %>% 
  filter(c1_c2_ratio == 2) %>% 
  mutate(beta_est_var = round(beta_est_var, 2)) %>%
  select(`G` = n_clusters, `# of Obs / Cluster (2)` = n_obs_per_cluster, `Var(beta_est) (2)` = beta_est_var)

r5_df_poisson <- res_poisson_opt %>% 
  filter(c1_c2_ratio == 5) %>% 
  mutate(beta_est_var = round(beta_est_var, 2)) %>%
  select(`# of Obs / Cluster (5)` = n_obs_per_cluster, `Var(beta_est) (5)` = beta_est_var)

r10_df_poisson <- res_poisson_opt %>% 
  filter(c1_c2_ratio == 10) %>% 
  mutate(beta_est_var = round(beta_est_var, 2)) %>%
  select(`# of Obs / Cluster (10)` = n_obs_per_cluster, `Var(beta_est) (10)` = beta_est_var)

r20_df_poisson <- res_poisson_opt %>% 
  filter(c1_c2_ratio == 20) %>% 
  mutate(beta_est_var = round(beta_est_var, 2)) %>%
  select(`# of Obs / Cluster (20)` = n_obs_per_cluster, `Var(beta_est) (20)` = beta_est_var)

# Combined Table
combined_df_poisson <- bind_cols(r2_df_poisson, r5_df_poisson, r10_df_poisson, r20_df_poisson)

stacked_combined_df <- bind_rows(combined_df, combined_df_poisson) 

stacked_combined_df %>%
  gt() %>%
  tab_row_group(
    group = "Poisson",
    rows = 10:18
  ) %>%
    tab_row_group(
    group = "Normal",
    rows = 1:9
  ) %>%
  tab_spanner(
    label = "c1/c2=2",
    columns = c(`# of Obs / Cluster (2)`, `Var(beta_est) (2)`)
  ) %>%
  tab_spanner(
    label = "c1/c2=5",
    columns = c(`# of Obs / Cluster (5)`, `Var(beta_est) (5)`)
  ) %>%
  tab_spanner(
    label = "c1/c2=10",
    columns = c(`# of Obs / Cluster (10)`, `Var(beta_est) (10)`)
  ) %>%
  tab_spanner(
    label = "c1/c2=20",
    columns = c(`# of Obs / Cluster (20)`, `Var(beta_est) (20)`)
  ) %>%
  cols_label(
    `# of Obs / Cluster (2)` = "R",
    `# of Obs / Cluster (5)` = "R",
    `# of Obs / Cluster (10)` = "R",
    `# of Obs / Cluster (20)` = "R",
    `Var(beta_est) (2)` = "Var(beta_hat)",
    `Var(beta_est) (5)` = "Var(beta_hat)",
    `Var(beta_est) (10)` = "Var(beta_hat)",
    `Var(beta_est) (20)` = "Var(beta_hat)"
  ) %>%
  tab_options(
    table.font.size = px(8),
    heading.title.font.size = px(8)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold", align = "center"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#c9ecb4")),
    locations = cells_body(columns = c("G","# of Obs / Cluster (2)", "Var(beta_est) (2)",

                                       "# of Obs / Cluster (20)", "Var(beta_est) (20)"), rows = 9)
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#c9ecb4")),
    locations = cells_body(columns = c("G","# of Obs / Cluster (10)", "Var(beta_est) (10)",
                                       "# of Obs / Cluster (5)", "Var(beta_est) (5)"), rows = 8)
  ) %>%
  tab_style(
    style = list(
      cell_fill(color = "#c9ecb4")),
    locations = cells_body(columns = c("G",
                                       "# of Obs / Cluster (5)", "Var(beta_est) (5)",
                                       "# of Obs / Cluster (20)", "Var(beta_est) (20)"), rows =18)) %>%
  tab_style(
    style = list(
      cell_fill(color = "#c9ecb4")),
    locations = cells_body(columns = c("G","# of Obs / Cluster (2)", "Var(beta_est) (2)",
                                       "# of Obs / Cluster (10)", "Var(beta_est) (10)"), rows = 17)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold", align = "center"), # Bold the spanner text
    locations = cells_column_spanners(everything())
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"), # Bold the row group labels
    locations = cells_row_groups()
  ) %>%
  tab_header(
    title = "Table 1: Simulation Results Vary by Number of Clusters and Relative Costs"
  )
```

### Normal Distribution

Under the Normal distribution, `Figure 1` visualizes the relationship between the variance of the treatment effect estimate ($\hat{\beta}$) and the number of clusters ($G$), stratified by relative cost ratios ($c_1/c_2 = 2, 5, 10, 20$). Each line represents a specific cost ratio, and the labels indicate the corresponding number of measurements per cluster ($R$) at different cluster sizes. As the number of clusters increases, the variance of the treatment effect estimate decreases, which is consistent with the results presented in `Table 1`. However, this decreasing trend follows a pattern of diminishing returns. Beyond a certain threshold, around $G = 40$, adding more clusters results in only marginal reductions in variance, suggesting that there is an optimal point beyond which additional clusters do not significantly improve precision. This indicates that while increasing the number of clusters improves the precision of treatment effect estimates, there is a limit to the benefit gained from simply adding more clusters.

`Figure 1` also shows the impact of varying cost ratios on resource allocation between the number of clusters and the number of measurements per cluster. Higher cost ratios lead to a lower relative cost of additional measurements ($c_2$), allowing for more resources to be allocated to increasing the number of measurements per cluster. Interestingly, when $c_1/c_2 = 10$, the decreasing trend in variance becomes flatter, indicating that the variance appears to be smaller overall compared to the other cost ratios. This suggests that for higher cost ratios, the allocation of resources to more measurements per cluster is more effective at reducing variance. However, as the cost ratios increases further to 20, meaning there is a large number of measurements per cluster, the allocation of more resources to measurements does not necessarily result in lower variance on average. Thus, a higher number of measurements per cluster (as seen in $c_1/c_2 = 20$) does not always guarantee better precision, highlighting the importance of the cost structure in determining the optimal balance between the number of clusters and the number of measurements per cluster.

```{r warning=FALSE, message=FALSE, out.width="80%", fig.align='center'}
# Assign colors for the c1/c2 ratio
colors_normal <- c(
  "2" = brewer.pal(9, "YlOrRd")[4],  
  "5" = brewer.pal(9, "YlOrRd")[6],  
  "10" = brewer.pal(9, "YlOrRd")[8],  
  "20" = brewer.pal(9, "YlOrRd")[9]
)


# Plot with adjusted text position
ggplot(res_normal_opt, aes(x = n_clusters, y = beta_est_var)) +
  geom_line(aes(color = as.factor(c1_c2_ratio)), size = 0.8) +
  geom_point(aes(color = as.factor(c1_c2_ratio)), size = 1) +
  geom_text(aes(label = n_obs_per_cluster, color = as.factor(c1_c2_ratio)),
            position = position_nudge(y = 0.025, x=2.2),  # Nudging text upward
            size = 2, check_overlap = TRUE) +
  scale_color_manual(values = colors_normal, name = "Relative costs (c1/c2)") +
  facet_wrap(~c1_c2_ratio, nrow = 1) +
  labs(
    title = "Figure 1: Variance of Beta Estimates vs. Number of Clusters (Normal)",
    x = "Number of Clusters",
    y = "Variance of Beta Estimates"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8),
    legend.position = "bottom"
  )
```

### Poisson

Under the Poisson distribution, the variance of $\beta_{est}$ follows a similar trend but is generally lower compared to the Normal distribution, particularly for smaller G. For instance, at G=10 and $\frac{c_1}{c_2}$=2, the variance is 0.29 under Poisson compared to 0.44
under Normal. Similar to the Normal distribution, increasing the number of clusters ($G$) results in a reduction in variance, but diminishing returns are observed. After a certain threshold, around $G = 40$, the variance continues to decrease at a slower rate, suggesting an optimal point beyond which adding more clusters does not substantially reduce variance. The plot also reveals the impact of varying cost ratios ($c_1/c_2$) on the allocation of resources between the number of clusters and the number of measurements per cluster ($R$). For lower cost ratios ($c_1/c_2 = 2$) and higher cost ratios ($c_1/c_2 = 20$), the overall variance of $\hat{\beta}$ is smaller compared to the moderate cost ratios of $c_1/c_2 = 5$ and $c_1/c_2 = 10$.


```{r warning=FALSE, message=FALSE, out.width="80%", fig.align='center'}
# Assign colors for the c1/c2 ratio
colors_poisson <- c(
  "2" = brewer.pal(9, "Blues")[5],  # Moderate orange
  "5" = brewer.pal(9, "Blues")[7],  # Deeper orange-red
  "10" = brewer.pal(9, "Blues")[8],  # Dark red
  "20" = brewer.pal(9, "Blues")[9]
)


# Plot with adjusted text position
ggplot(res_poisson_opt, aes(x = n_clusters, y = beta_est_var)) +
  geom_line(aes(color = as.factor(c1_c2_ratio)), size = 0.8) +
  geom_point(aes(color = as.factor(c1_c2_ratio)), size = 1) +
  geom_text(aes(label = n_obs_per_cluster, color = as.factor(c1_c2_ratio)),
            position = position_nudge(y = 0.025, x=2.2),  # Nudging text upward
            size = 2, check_overlap = TRUE) +
  scale_color_manual(values = colors_poisson, name = "Relative costs (c1/c2)") +
  facet_wrap(~c1_c2_ratio, nrow = 1) +
  labs(
    title = "Figure 2: Variance of Beta Estimates vs. Number of Clusters (Poisson)",
    x = "Number of Clusters",
    y = "Variance of Beta Estimates"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8),
    legend.position = "bottom"
  )
```

In summary, the results demonstrate that increasing number of clusters reduces the
variance of $\beta_{est}$ under both distributions, with more pronounced
improvements at smaller G values. Lower cost ratios
($\frac{c_1}{c_2}$=2 or 5) result in higher variances due to fewer
measurements per cluster. 


## Varying Data Generation

```{r warning=FALSE, message=FALSE}
# Function to simulate for different sigma2 and beta under optimal # of clusters and # of obs/cluster
sim_normal_vary <- function(normal_opt, sigma2_values, beta_values, B, c1, alpha, gamma2, n_sim, alpha_level = 0.05) {
  
  # initialize storage for all results
  all_results <- list()
  
  # Loop over within-cluster variacen and treatment effect
  for (row in seq_len(nrow(normal_opt))) {
    n_clusters <- normal_opt$n_clusters[row]
    n_obs_per_cluster <- normal_opt$n_obs_per_cluster[row]
    c1_c2_ratio <- normal_opt$c1_c2_ratio[row]
    
    for (sigma2 in sigma2_values) {
      for (beta in beta_values) {
        # Run simulation for current parameter set
        result <- sim_normal(
          n_clusters = n_clusters,
          B = B,
          c1 = c1,
          c1_c2_ratio = c1_c2_ratio,
          alpha = alpha, 
          beta = beta,
          gamma2 = gamma2,
          sigma2 = sigma2,
          n_sim = n_sim,
          alpha_level = alpha_level
        )
        
        all_results <- append(all_results, list(
          result$metrics %>% mutate(sigma2 = sigma2, beta = beta, c1_c2_ratio = c1_c2_ratio)
        ))
      }
    }
  }
  
  # Combine all results into a single data frame
  combined_results <- bind_rows(all_results)
  return(combined_results)
}

# extract optimal combination of number of cluster and number of obs / cluster
normal_opt <- res_normal_opt %>%
  group_by(c1_c2_ratio) %>%
  summarize(
    n_clusters = n_clusters[which.min(beta_est_var)],  # Optimal number of clusters
    n_obs_per_cluster = n_obs_per_cluster[which.min(beta_est_var)]
  )

# Set parameters
sigma2_values <- seq(0.1, 10, length.out=10)
beta_values <- c(0.05, 0.5, 1.5)
B <- 2000
c1 <- 20
alpha <- 2
gamma2 <- 1
n_sim <- 100

# Run simulations
# res_normal_vary <- sim_normal_vary(
#   normal_opt = normal_opt,
#   sigma2_values = sigma2_values,
#   beta_values = beta_values,
#   B = B,
#   c1 = c1,
#   alpha = alpha,
#   gamma2 = gamma2,
#   n_sim = n_sim
# )
# 
# write.csv(res_normal_vary, "res_normal_vary.csv")
res_normal_vary <- read.csv("res_normal_vary.csv")
```

```{r warning=FALSE, message=FALSE}
# Function to simulate for different sigma2 and beta under optimal # of clusters and # of obs/cluster
sim_poisson_vary <- function(poisson_opt, gamma2_values, beta_values, B, c1, alpha, n_sim, alpha_level) {
  all_results <- list()
  
  # Loop over between-cluster variance and treatment effect size
  for (row in seq_len(nrow(poisson_opt))) {
    n_clusters <- poisson_opt$n_clusters[row]
    n_obs_per_cluster <- poisson_opt$n_obs_per_cluster[row]
    c1_c2_ratio <- poisson_opt$c1_c2_ratio[row]
    
    for (gamma2 in gamma2_values) {
      for (beta in beta_values) {
        result <- sim_poisson(
          n_clusters = n_clusters,
          B = B,
          c1 = c1,
          c1_c2_ratio = c1_c2_ratio,
          alpha = alpha,
          beta = beta,
          gamma2 = gamma2,
          n_sim = n_sim,
          alpha_level = alpha_level
        )
        
        all_results <- append(all_results, list(
          result$metrics %>% mutate(gamma2 = gamma2, beta = beta, c1_c2_ratio = c1_c2_ratio)
        ))
      }
    }
  }
  
  combined_results <- bind_rows(all_results)
  return(combined_results)
}

# Set parameters
 beta_values <- c(0.05, 0.5, 1.5)
# beta_values <- c(0.5, 1.5, 3)
gamma2_values <- seq(0.1, 10, length.out = 10)

# Extract optimal configurations from varying_results
poisson_opt <- res_poisson_opt %>%
  group_by(c1_c2_ratio) %>%
  summarize(
    n_clusters = n_clusters[which.min(beta_est_var)],  # Optimal number of clusters
    n_obs_per_cluster = n_obs_per_cluster[which.min(beta_est_var)]
  )

# Run the simulation
# res_poisson_vary <- sim_poisson_vary(
#   poisson_opt = poisson_opt,
#   gamma2_values = gamma2_values,
#   beta_values = beta_values,
#   B = 2000,
#   c1 = 20,
#   alpha = alpha,
#   n_sim = 100,
#   alpha_level = 0.05
# )
# 
# write.csv(res_poisson_vary, "res_poisson_vary.csv")
res_poisson_vary <- read.csv("res_poisson_vary.csv")
```

`Table 2` summarizes simulation results, highlighting the variance of the treatment effect estimate ($\hat{\beta}$) and power across varying levels of variance ($\sigma^2$ for Normal, $\gamma^2$ for Poisson), treatment effect sizes ($\beta$), and relative cost ratios ($c_1/c_2 = 2, 5, 10, 20$). For the Normal distribution, higher $\sigma^2$ results in increased variance of $\hat{\beta}$ and reduced power, particularly for smaller treatment effects ($\beta = 0.05$), while larger effects ($\beta = 1.5$) consistently yield lower variances and higher power. Similarly, for Poisson outcomes, as $\gamma^2$ increases, both the variance of $\hat{\beta}$ and the loss of power become more pronounced for smaller $\beta$. Larger treatment effects ($\beta = 1.5$) are associated with lower variances and higher power in both distributions. The optimal number of clusters ($G$) and measurements per cluster ($R$), identified in prior analyses, reflect a trade-off between within-cluster and between-cluster variability, influenced by the cost ratio. Higher cost ratios ($c_1/c_2 = 20$) facilitate more measurements per cluster but do not consistently minimize variance or maximize power, underscoring the need to carefully balance variability, effect size, and cost considerations when selecting optimal study designs.

```{r , out.width="80%"}
# Results Table (Normal)
df_normal_vary <- res_normal_vary %>% 
  arrange(c1_c2_ratio, sigma2, beta) %>% 
  filter( sigma2 %in% c(0.1, 4.5, 10)) %>% 
  mutate(beta_est_var = round(beta_est_var, 2),
         `G (R) (n)`=paste(`n_clusters`, "(", `n_obs_per_cluster`, ")")) %>%
  select(c1_c2_ratio, `sigma^2 (gamma^2)`= sigma2, beta, `G (R) (n)`, `Var(beta_est) (n)` = beta_est_var, `power (n)` = power) 

# Results Table (Poisson)
df_poisson_vary <- res_poisson_vary %>% 
  arrange(c1_c2_ratio, gamma2, beta) %>% 
  filter(gamma2 %in% c(0.1, 4.5, 10)) %>% 
  mutate(beta_est_var = round(beta_est_var, 2),
         `G (R) (p)`=paste(`n_clusters`, "(", `n_obs_per_cluster`, ")")) %>%
  select(`G (R) (p)`,`Var(beta_est) (p)` = beta_est_var, `power (p)` = power) 

df_com_vary <- bind_cols(df_normal_vary, df_poisson_vary)
df_com_vary %>%
  gt() %>%
  tab_spanner(
    label = "Normal",
    columns = c(`G (R) (n)`, `Var(beta_est) (n)`, `power (n)`)
  ) %>%
  tab_spanner(
    label = "Poisson",
    columns = c(`G (R) (p)`, `Var(beta_est) (p)`, `power (p)`)
  ) %>%
  cols_label(
    c1_c2_ratio = "c1/c2",
    #`sigma^2 (gamma^2)` = md("&sigma;<sup>2</sup> (&gamma;<sup>2</sup>)"),
    #beta = md("&beta;"),
    `G (R) (n)` = "G (R)",
    `G (R) (p)` = "G (R)",
    # `Var(beta_est) (n)` = md("Var(&beta;<sub>est</sub>)"),
    # `Var(beta_est) (p)` = md("Var(&beta;<sub>est</sub>)"),
    `Var(beta_est) (n)` = "Var(beta_hat)",
    `Var(beta_est) (p)` = "Var(beta_hat)",
    `power (n)` = "power",
    `power (p)` = "power"
  ) %>%
  tab_options(
    table.font.size = px(8),
    heading.title.font.size = px(8)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold", align = "center"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = cell_text(weight = "bold", align = "center"), # Bold the spanner text
    locations = cells_column_spanners(everything())
  ) %>%
  tab_header(
    title = "Table 2: Simulation Results Vary by Variance and Treatment Effect"
  )
```

### Normal

`Figure 3` illustrates the relationship between the variance of the treatment effect estimate ($\hat{\beta}$) and power, across varying values of within-cluster variance ($\sigma^2$) and relative cost ratios ($c_1/c_2 = 2, 5, 10, 20$). In the left panel, the variance of $\hat{\beta}$ increases as $\sigma^2$ grows, with higher values of $\sigma^2$ leading to less precise estimates of $\beta$. For smaller treatment effects ($\beta = 0.05$), the variance grows sharply, particularly when $\sigma^2$ exceeds 5. For larger treatment effects ($\beta = 1.5$), the variance remains relatively stable despite increases in $\sigma^2$, reflecting that larger effect sizes are more detectable even in the presence of residual variability. In the right panel, as expected, power decreases with increasing $\sigma^2$, especially for smaller treatment effects. In particular, lower cost ratios achieve slightly higher power for smaller values of $\sigma^2$, as more resources are allocated to within-cluster measurements, improving detection. However, for higher cost ratios, the emphasis shifts towards increasing the number of clusters, which leads to fewer measurements per cluster and can reduce power. For larger treatment effects, power remains relatively stable despite increases in $\sigma^2$, showing that larger treatment effects are more easily detected, even with increasing within-cluster variance.

In this study, clusters are represented by individuals, and the lower level corresponds to the number of repeated measurements ($R$) per individual. The intra-class correlation coefficient (ICC) quantifies the proportion of total variance attributable to variability between individuals (clusters). As $\sigma^2$, the within-individual variance, increases under a fixed between-individual variance ($\gamma^2$), the ICC decreases. When $\text{ICC} \approx 1$, repeated measurements within an individual are perfectly correlated, so additional measurements provide no new information, and the effective sample size reduces to the number of individuals (clusters). Conversely, when $\text{ICC} = 0$, measurements within an individual behave independently, and the available sample size is the total number of repeated measurements, maximizing the informational value of the study. As $\sigma^2$ increases and ICC decreases, the clusters contribute less unique information, making within-cluster data less informative and increasing the variance of $\hat{\beta}$. This diminished efficiency highlights the critical role of ICC in study design, as higher ICC values reduce the effective sample size and inflate the variance of the treatment effect estimate. These findings underscore the importance of carefully considering ICC and its relationship to within-cluster and between-cluster variability when designing cluster randomized trials.

```{r, fig.align='center'}
final_results_long_normal <- res_normal_vary %>%
  pivot_longer(cols = c(beta_est_var, power, ci_coverage), 
               names_to = "metric", 
               values_to = "value")

# Define y-axis limits for each metric
y_lim_normal <- list(
  power = c(0, 1),
  beta_est_var = c(0, max(res_normal_vary$beta_est_var, na.rm = TRUE)),
  avg_ci_coverage = c(0.8, 1)
)

colors_normal <- c(
  "2" = brewer.pal(9, "YlOrRd")[5],  # Moderate orange
  "5" = brewer.pal(9, "YlOrRd")[7],  # Deeper orange-red
  "10" = brewer.pal(9, "YlOrRd")[8],  # Dark red
  "20" = brewer.pal(9, "YlOrRd")[9]
)

# Updated function to create a plot for a given metric
plot_metric_consistent_normal <- function(metric, metric_label) {
  ggplot(final_results_long_normal %>% filter(metric == !!metric), 
         aes(x = sigma2, y = value, color = as.factor(c1_c2_ratio))) +
    geom_line(size = 0.3, alpha=0.8) +
    geom_point(size = 0.6) +
    facet_wrap(~ beta, scales = "fixed") +
    scale_color_manual(values = colors_normal,
                       name = "c1/c2 Ratio") +
    coord_cartesian(ylim = y_lim_normal[[metric]]) +
    labs(
      #title = paste(metric_label, "by Gamma^2, Colored by c1/c2 Ratio"),
      x = expression(sigma^2),
      y = metric_label
    ) +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 6),
      axis.title = element_text(size = 8, face = "bold"),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      legend.position = "bottom"
    )
}


# Generate plots for each metric
plot_beta_est_var_normal <- plot_metric_consistent_normal("beta_est_var", "Variance of Beta Estimate")
plot_power_normal <- plot_metric_consistent_normal("power", "Power")

# Display the plots (one at a time)
combined_plot_normal <- ggarrange(plot_beta_est_var_normal, plot_power_normal, common.legend = TRUE, # Use a common legend
  legend = "bottom")

final_plot_normal <- annotate_figure(
  combined_plot_normal,
  top = text_grob("Figure 3: Variance of Beta Estimate and Power Across c1/c2 Ratios (Normal)", size = 10, face = "bold")
)
final_plot_normal
```

### Poisson

`Figure 4` illustrates the relationship between the variance of the treatment effect estimate ($\hat{\beta}$) and power, across varying values of between-cluster variance ($\gamma^2$) and relative cost ratios ($c_1/c_2 = 2, 5, 10, 20$). In the left panel, the variance of $\hat{\beta}$ increases as $\gamma^2$ grows, with higher values of $\gamma^2$ corresponding to increased between-cluster heterogeneity. This reflects the greater variability introduced by differences across clusters, reducing the precision of the treatment effect estimate. For smaller treatment effects ($\beta = 0.05$), the variance increases more sharply with $\gamma^2$, indicating that small effect sizes are more sensitive to between-cluster variability. For larger treatment effects ($\beta = 1.5$), the variance grows more gradually, highlighting that larger effect sizes are less affected by increased between-cluster variance.

In the Poisson setting, $\gamma^2$ directly influences the intra-class correlation coefficient (ICC), which quantifies the proportion of total variance attributable to differences between clusters. As $\gamma^2$ increases, ICC rises, meaning the measurements within a cluster become more similar to each other. This reduces the effective sample size, as adding more measurements within a highly correlated cluster contributes less unique information, thereby inflating the variance of $\hat{\beta}$. When ICC is low (small $\gamma^2$), clusters contribute more unique information, leading to smaller variances in $\hat{\beta}$.

In the right panel, power decreases as $\gamma^2$ increases, particularly for smaller treatment effects ($\beta = 0.05$), where detecting the effect becomes more difficult due to greater between-cluster variability. Lower cost ratios ($c_1/c_2 = 2$) achieve slightly higher power for smaller $\gamma^2$, as they allocate more resources to within-cluster measurements, improving the ability to detect effects. Higher cost ratios ($c_1/c_2 = 10, 20$), which prioritize the number of clusters, maintain more stable power across larger $\gamma^2$, highlighting a trade-off between cluster count and measurement allocation in optimizing power. For large treatment effects ($\beta = 1.5$), power remains relatively robust to changes in $\gamma^2$, underscoring the strong signal associated with larger effects.

These findings emphasize the interplay between between-cluster variability, ICC, and study design parameters. In Poisson-distributed outcomes, higher ICC (driven by larger $\gamma^2$) increases the variance of $\hat{\beta}$, reduces power for smaller treatment effects, and underscores the importance of carefully balancing cluster count and within-cluster measurements in study design.

```{r, fig.align='center'}
# Step 4: Visualize results
final_results_long_poisson <- res_poisson_vary %>%
  pivot_longer(cols = c(beta_est_var, power, ci_coverage), 
               names_to = "metric", 
               values_to = "value")

# Define consistent y-limits for each metric
y_lim_poisson <- list(
  beta_est_var = c(0, max(res_poisson_vary$beta_est_var, na.rm = TRUE)),
  power = c(0, 1),
  ci_coverage = c(0.8, 1)
)

colors_poisson <- c(
  "2" = brewer.pal(9, "Blues")[5],  # Moderate orange
  "5" = brewer.pal(9, "Blues")[7],  # Deeper orange-red
  "10" = brewer.pal(9, "Blues")[8],  # Dark red
  "20" = brewer.pal(9, "Blues")[9]
)

plot_metric_consistent_y <- function(metric, metric_label) {
  ggplot(final_results_long_poisson %>% filter(metric == !!metric), 
         aes(x = gamma2, y = value, color = as.factor(c1_c2_ratio))) +
    geom_line(size = 0.3, alpha=0.8) +
    geom_point(size = 0.6) +
    facet_wrap(~ beta, scales = "fixed") +
    scale_color_manual(values = colors_poisson,
                       name = "c1/c2 Ratio") +
    coord_cartesian(ylim = y_lim_poisson[[metric]]) +
    labs(
      #title = paste(metric_label, "by Gamma^2, Colored by c1/c2 Ratio"),
      x = expression(gamma^2),
      y = metric_label
    ) +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 6),
      axis.title = element_text(size = 8, face = "bold"),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      legend.position = "bottom"
    )
}

# Plot beta_est_var
plot_beta_est_var_poisson <- plot_metric_consistent_y("beta_est_var", "Variance of Beta Estimate")
plot_power_poisson <- plot_metric_consistent_y("power", "Power")

# Display the plots (one at a time)
combined_plot_poisson <- ggarrange(plot_beta_est_var_poisson, plot_power_poisson, common.legend = TRUE, # Use a common legend
  legend = "bottom")

final_plot_poisson <- annotate_figure(
  combined_plot_poisson,
  top = text_grob("Figure 4: Variance of Beta Estimate and Power Across c1/c2 Ratios (Poisson)", size = 10, face = "bold")
)
final_plot_poisson
```


Both the Normal and Poisson models show that increasing variance components ($\sigma^2$ in Normal, $\gamma^2$ in Poisson) reduces the precision of the treatment effect estimate ($\hat{\beta}$), highlighting the critical role of intra-class correlation (ICC) in study design. Lower ICC values, which imply less correlation within clusters, increase the effective sample size and reduce the variance of $\hat{\beta}$. However, the mechanisms differ due to the underlying data structures. In the Normal model, $\sigma^2$ represents residual variability independent of cluster means, causing a linear increase in $\hat{\beta}$'s variance as noise grows. In contrast, the Poisson models $\gamma^2$ affects the variability in the log-scale parameter $\log(\mu_i)$, which indirectly influences both within- and between-cluster variability by broadening the expected count range. The Normal model assumes additive symmetric residuals, while the Poisson model deals with skewed, non-negative count data, leading to different sensitivities to parameter changes. These differences underscore the importance of carefully considering the data structure and ICC when designing studies, as they shape the balance between within-cluster and between-cluster variability and their effects on $\hat{\beta}$'s precision.



# Discussion
This study investigates optimal design strategies for cluster randomized trials under budgetary constraints by systematically examining trade-offs between within-cluster and between-cluster variability. The results show that increasing the number of clusters consistently reduces the variance of $\hat{\beta}$ across all cost ratio levels, though with diminishing returns beyond a certain point.

For Normal outcomes, increasing $\sigma^2$ weakens clustering effects by reducing the intra-class correlation (ICC), thereby increasing the variability of $\hat{\beta}$ and diminishing precision. As $\sigma^2$ grows, power decreases significantly for smaller treatment effects but remains relatively robust for larger effects ($\beta = 1.5$). In contrast, for Poisson outcomes, $\gamma^2$ strongly influences ICC, with higher $\gamma^2$ values resulting in greater between-cluster variability and inflating $\hat{\beta}$'s variance. The impact of $\gamma^2$ is particularly pronounced for smaller treatment effects, where power declines sharply as ICC rises.

Interestingly, while larger treatment effects consistently yield lower variances and higher power across both distributions, the sensitivity to variance components ($\sigma^2$ and $\gamma^2$) and cost ratio adjustments varies between Normal and Poisson outcomes. For Poisson-distributed data, increasing $\gamma^2$ directly increases ICC, diminishing the effective sample size and reducing the information gained from additional measurements within clusters. This difference reflects the distinct hierarchical data structures, with Normal outcomes exhibiting additive residual variability and Poisson outcomes demonstrating multiplicative, count-based heterogeneity.

These findings highlight the importance of tailoring study designs to the underlying data-generating mechanisms and variance structures. For example, in scenarios with high between-cluster variability (e.g., high $\gamma^2$), designs with more clusters but fewer measurements per cluster may be preferable. Conversely, when within-cluster variability dominates (e.g., high $\sigma^2$), allocating resources toward within-cluster measurements could improve precision and power. The nuanced interplay between ICC, variance components, and cost ratios emphasizes the need for careful planning and consideration of the data structure to achieve cost-efficient and statistically robust trial designs.



# Limitation

This project has several limitations. First, the number of simulations conducted for each scenario was limited to 100 due to computational constraints. While this choice was made to balance running time and feasibility, it may not fully capture the variability in the estimates, potentially leading to less stable conclusions about the optimal combinations of clusters ($G$) and measurements per cluster ($R$). Increasing the number of simulations would provide more robust and reliable results, albeit at the cost of significantly greater computational resources.

Second, the sequential approach used to determine the optimal design parameters introduces a limitation. The number of clusters ($G$) and relative cost ratios ($c_1/c_2$) were varied first to identify the optimal combination of $G$ and $R$, and then the variance components ($\sigma^2$, $\gamma^2$) and treatment effects ($\beta$) were varied while keeping the optimal $G$ and $R$ fixed. This approach does not allow for the simultaneous variation of $G$, $R$, variance components, and treatment effects. As a result, potential interactions between these parameters could not be fully explored, which may limit the generalizability of the findings. Future work should consider varying all parameters together to better capture their joint effects on study outcomes.

Finally, the exploration of intra-class correlation (ICC) under the Normal distribution focused on varying $\sigma^2$ while keeping other parameters fixed. Although this approach provided insights into how ICC impacts the variance of $\hat{\beta}$, $\gamma^2$ was not varied. Both $\sigma^2$ and $\gamma^2$ contribute to ICC, but their individual impacts on the variance of $\hat{\beta}$ may differ. By not varying $\gamma^2$, the analysis may overlook important differences in how these parameters affect ICC and the resulting variance of $\hat{\beta}$. Future studies should address this by simultaneously varying $\sigma^2$ and $\gamma^2$ to provide a more comprehensive understanding of their respective roles in influencing ICC and study precision.



# References
Clancy, M., et al. (2020). Evaluating the effectiveness of a hand hygiene protocol in hospitals: A cluster randomized trial. Journal of Hospital Infection, 104(3), 241248.

Karimian, Z., et al. (2023). Effects of a dietary intervention on glycemic control with repeated measurements: A cluster-based study. Cardiovascular Diabetology, 22(4), 300312.


\newpage

# Code Appendix

```{r ref.label = knitr::all_labels()}
#| echo: true
#| eval: false
```
